# Tools {#tools}

This chapter covers various statistical and computational tools that will be used in subsequent chapters, collected here for easy reference.

## Probability distributions

  * cdf
  * density
  * quantile
  * Normal
  * t
  * Chi-squared
  * Gamma

## Central limit theorem

## Pairwise distances

Many anomaly detection algorithms are based on pairwise distances between observations. If there are $n$ observations, then there are $n(n-1)/2$ pairwise distances to compute, so this is an $O(n^2)$ operation which can take a very long time for large $n$.

Suppose our observations are denoted by $\bm{y}_1,\dots,\bm{y}_n$. For now, we will assume these are numerical vectors of observations. Later in the book, we will allow the observations to be more complicated, where they may denote images, or video, or probability distributions, for example. But for the first half of the book, we will only deal with numerical data.

When we have numerical data, we will usually want to use a Euclidean distance, named after the famous Greek mathematician Euclid. That is, the distance between points $\bm{y}_i = (y_{i1},\dots,y_{id})'$ and $\bm{y}_j=(y_{j1},\dots,y_{jd})'$ is given by
$$\|\bm{y}_i - \bm{y}_j\| = \sqrt{\sum_{k=1}^d (y_{ik}-y_{jk})^2}.$$
This is also known as the $L_2$ distance.

For $d=1$ or $d=2$, this is the physical distance between the points when plotted on a strip plot or a scatterplot (provided there is no jittering used).

The `dist()` function will return a distance matrix containing all pairwise distances computed in this way. Here is an example using only the first five observations of the `old_faithful` data set (omitting the time stamp). Because the distances are symmetric, only the lower triangle of the matrix is computed.

```{r}
oldfaithful %>%
  select(-time) %>%
  head(5) %>%
  dist()
```

When the variables have very different scales, the variables with the largest ranges will dominate the distance measures. In this example, durations are much longer than waiting times, and so the duration variable is dominating the calculation of neighbours. Consequently, it is often preferable to scale the data before computing distances.

```{r}
oldfaithful %>%
  select(-time) %>%
  scale() %>%
  head(5) %>%
  dist()
```

The `scale()` function subtracts the mean and divides by the standard deviation for each column of data. Therefore, all variables in the resulting scaled data have mean zero and standard deviation equal to 1.

The `dist()` function will also compute other types of distances which may be more appropriate for some kinds of data. These are specified using the `method` argument and defined below.

Manhattan:
: Absolute distance between the two vectors:
$$\|\bm{y}_i - \bm{y}_j\| = \sum_{k=1}^d |y_{ik}-y_{jk}|.$$
This is also known as the $L_1$ distance.  It is called the Manhattan distance as it gives the shortest path  between the corners of city blocks (denoted as points on a grid) when those blocks are rectangular, as they mostly are in Manhattan. For the same reason, it is also sometimes called the "taxicab" distance or the "city block" distance.

Minkowski:
: This generalizes the Manhattan and Euclidean distances to use powers of $p$ to define the $L_p$ distance:
$$\|\bm{y}_i - \bm{y}_j\| = \left(\sum_{k=1}^d (y_{ik}-y_{jk})^p\right)^{1/p}.$$
The value of $p$ is specified using the `p` argument. It is named after the German mathematician Hermann Minkowski.

Maximum:
: Maximum distance between any components of $\bm{y}_i$ and $\bm{y}_j$:
$$\|\bm{y}_i - \bm{y}_j\| = \max_k |y_{ik}-y_{jk}|.$$

Canberra:
: $$\|\bm{y}_i - \bm{y}_j\| = \sum_{k=1}^d \frac{|y_{ik}-y_{jk}|}{|y_{ik}|+|y_{jk}|}.$$
Terms with zero numerator and denominator are omitted from the sum. This distance was introduced by two Australian scientists, Godfrey Lance and Bill Williams, who named it after their home city of Canberra. It scales the distances between components by the size of the components, so there is an inbuilt scaling which avoids problems with variables on different units. However, if there are components which are close to zero, but not identical to zero, the measure is numerically unstable. So it is better to scale the data explicitly before computing a distance, unless the variables cannot take small values.

Binary:
: This is designed for binary vectors where each element is either 0 or 1. The vectors are regarded as binary bits, so non-zero elements are ‘on’ and zero elements are ‘off’. The distance is the proportion of bits in which only one is on amongst those in which at least one is on. This may be useful for distances between logical variables.

## Nearest neighbours

Some algorithms only compute the pairwise distances of the $k$ nearest observations, although finding those observations requires some additional distances to be computed. For some types of distances, efficient solutions are available using kd trees [@Bentley1975;@Arya1998] that find the $k$ nearest neighbours to each observation in $O(n\log(n))$ time.

The calculation of $k$ nearest neighbours is useful for more than anomaly detection problems. It is also the basis of a popular classification method due to the Berkeley statisticians Evelyn Fix and Joe Hodges [@knn] which is often known as the "kNN algorithm".

Suppose we use the Old Faithful data to find eruptions that are neighbours in the (duration, waiting) space. The `dbscan` package uses kd trees to quickly identify the $k$ nearest observations to each eruption. As noted earlier, we will scale the data before computing any distances.

```{r}
# Find 5 nearest neighbours to each eruption
knn <- oldfaithful %>%
  select(duration, waiting) %>%
  scale() %>%
  dbscan::kNN(k = 5)
# First eruption in the data set
oldfaithful[1, ]
# Five closest observations
oldfaithful[knn$id[1, ], ]
```

For very large data sets, approximations are available which speed up the computation even more, but are less accurate in finding the $k$ nearest neighbours. The `approx` argument specifies a distance tolerance which makes the process faster for large data sets, although the neighbours returned may not be the exact nearest  neighbours.

```{r}
# Find 5 approximate nearest neighbours to each eruption
kann <- oldfaithful %>%
  select(duration, waiting) %>%
  scale() %>%
  dbscan::kNN(k = 5, approx = 2)
# Five closest observations
oldfaithful[kann$id[1, ], ]
```

Here the fifth closest observation has been omitted, and another nearby observation has been included instead, but otherwise the approximation has identified four of the five nearest observations.

## Kernel density estimation {#kde}

Kernel density estimation is the most popular method for nonparametric estimation of a probability density function.

Suppose we have $n$ univariate observations, $\{y_1,\dots,y_n\}$, which are independent draws from a probability distribution, and we want to estimate the underlying probability density function. The kernel estimate [see @WJ1995] is given by
\begin{equation}
  \hat{f}(y) = \frac{1}{n} \sum_{i=1}^n K_h(y-y_i),
  (\#eq:kde)
\end{equation}
where $K_h$ is a "kernel" function and $h$ is a bandwidth to be determined. We will use kernel functions that are themselves probability density functions with mean 0 and standard deviation $h$. For example, $K_h(u) = \exp(-u^2/h^2)/(h\sqrt{2\pi})$ is the Gaussian density function with mean zero and standard deviation $h$. Another popular kernel is the quadratic Epanechnikov kernel given by
$K_h(u) = [1-u^2/(5h^2)]_+ / (h4\sqrt{5}/3)$, where $x_+ = \max(x,0)$.

```{r of2021, include=FALSE}
of2021 <- oldfaithful %>%
  filter(as.Date(time) > "2021-01-01") %>%
  head(10) %>%
  mutate(eruption = row_number())
s <- sd(of2021$duration)
iqr <- IQR(of2021$duration)
h <- 0.9 * min(s, iqr/1.34) * NROW(of2021)^(-1/5)
```

Now we will apply \@ref(eq:kde) to the first ten Old Faithful eruption durations from 2021 that are in the `oldfaithful` data set. The kernel density estimate can be visualized as a sum of kernel functions centered over each observation with width determined by $h$, and height given by $K_h(0)/n$.

Let's suppose $K_h$ is a Gaussian density function and let $h = `r round(h, 1)`$ (I will explain this choice below). Then we get the following set of kernel functions.

```{r kde1, fig.asp=0.2, echo=FALSE, dependson="of2021"}
k <- tibble(x = seq(-3 * h, 3 * h, l = 1000)) %>%
  mutate(y = dnorm(x, 0, h) / 10)
of2021kde <- of2021 %>%
  mutate(k = list(k)) %>%
  unnest(k) %>%
  mutate(x = x + duration)
ggplot() +
  geom_line(data = of2021kde, aes(x = x, y = y, group = eruption), col = "gray") +
  geom_rug(
    data = of2021, mapping = aes(x = duration, y = 0),
    size = 1, length = unit(0.06, "npc"), sides = "b"
  ) +
  labs(x = "y = Duration (seconds)", y = latex2exp::TeX("$K_h(y - y_i)/n$"))
```

The vertical ticks show the location of the ten observations, while the grey lines show $\frac{1}{n}K_h(y - y_i)$ for $i=1,\dots,n$.

These functions are then added together, as in \@ref(eq:kde), to give the density estimate.

```{r kde2, fig.asp=0.5, echo=FALSE, dependson="kde1"}
ggplot() +
  geom_line(data = of2021kde, aes(x = x, y = y, group = eruption), col = "gray") +
  geom_rug(data = of2021, mapping = aes(x = duration, y = 0), sides = "b", size = 1) +
  geom_density(data = of2021, mapping = aes(x = duration), bw = h) +
  labs(x = "y = Duration (seconds)", y = "Density")
```

We made two choices when producing this estimate: the value of $h$ and the type of kernel $K_h$. If either was replaced with a different choice, the estimate would be different. For large data sets, it does not make much difference which kernel function is used.

The choice of $h$ is more difficult and will change the shape of the density estimate substantially. Here are three versions of the density estimate with bandwidths given by $h=5$, $h=15$ and $h=40$.

```{r kde3, fig.asp=0.5, echo=FALSE, dependson="of2021"}
of2021 %>%
  ggplot(aes(x = duration)) +
  geom_rug(sides = "b", size = 1) +
  geom_density(bw = 5, col = "#E69F00") +
  geom_density(bw = 15, col = "#56B4E9") +
  geom_density(bw = 40, col = "#009E73") +
  labs(x = "y = Duration (seconds)", y = "Density") +
  xlim(120, 320) +
  geom_label(x = 260, y = .02, label = "h = 5", col = "#E69F00") +
  geom_label(x = 275, y = .01, label = "h = 15", col = "#56B4E9") +
  geom_label(x = 320, y = .004, label = "h = 40", col = "#009E73")
```

When $h$ is too small, the density estimate is very rough with many peaks and troughs. When $h$ is too large, the density estimate is too smooth and we fail to see any features in the data. A popular choice for $h$ uses Silverman's "rule-of-thumb" [@Silverman1986,48]:
\begin{equation}
  h = 0.9 \min(s, \text{IQR}/1.34) n^{-1/5},
  (\#eq:ruleofthumb)
\end{equation}
where $s$ is the sample standard deviation and $\text{IQR} = \hat{q}_{0.75} - \hat{q}_{0.25}$. This tends to work reasonably well for most data sets. For the 10 observations in the example above, it gives $h = `r round(h, 1)`$, which is the value we used.

This example using only 10 observations was purely to illustrate the method. Let's now estimate a kernel density estimate for the full data set (other than that one pesky duration of 2 hours).

We will use the `geom_density()` function, which uses the Gaussian kernel and Silverman's rule-of-thumb by default, although other choices are available.

```{r oldfaithful4, fig.asp=0.45, fig.cap="Kernel density estimate of Old Faithful eruption durations since 2015, omitting the one eruption that lasted nearly two hours."}
oldfaithful %>%
  filter(duration < 7000) %>%
  ggplot(aes(x = duration)) +
  geom_density() +
  geom_rug() +
  labs(x = "Duration (seconds)")
```

```{r ofbw2, include=FALSE}
of <- oldfaithful %>%
  filter(duration < 7000)
s <- sd(of$duration)
iqr <- IQR(of$duration)
h <- 0.9 * min(s, iqr / 1.34) * NROW(of)^(-1 / 5)
```

As this is a much bigger data set (with `r NROW(of)` observations), the selected bandwidth is smaller and is now $h = `r round(h, 2)`$. Here the estimate has clearly identified the two groups of eruptions, one much larger than the other. The extreme observation of 1 second, and the unusual observations between 140 and 180 seconds are in the areas of low density. Later we will use density estimates at each observation to identify anomalous points.

### Statistical properties {-}

The statistical properties of the kernel density estimator \@ref(eq:kde) have been extensively studied, and are described in several books including @WJ1995 and @Scott2015.

An important asymptotic result is that the mean square error (MSE) of $\hat{f}(y)$ is
\begin{equation}
  \text{E}\left[(\hat{f}(y) - f(y))^2\right] \approx
  \frac{1}{4}h^4[f''(y)]^2 + \frac{f(y)R(K)}{nh} ,
  (\#eq:mse)
\end{equation}
where $R(K) = \int K^2(u)du$ is the "roughness" of the kernel function. An estimator is "consistent" if the MSE goes to zero as the sample size $n$ goes to infinity. So \@ref(eq:kde) gives a consistent estimator of the underlying density $f$ when both terms in \@ref(eq:mse) go to zero. That is,
\begin{equation}
  \lim_{n\rightarrow\infty} h = 0
  \qquad\text{and}\qquad
  \lim_{n\rightarrow\infty} nh = \infty,
  (\#eq:asymptotickde)
\end{equation}
and so $h$ should decrease slowly as $n$ increases. Note that these conditions hold for \@ref(eq:ruleofthumb).

If we integrate the MSE given by \@ref(eq:mse) over $y$ (assuming $f$ is sufficiently smooth for the integral to exist), we obtain the mean integrated squared error (MISE) given by
\begin{equation}
  \text{E}\int \left[(\hat{f}(y) - f(y))^2\right] dy \approx
  \frac{1}{4}h^4R(f'') + \frac{R(K)}{nh} ,
  (\#eq:mise)
\end{equation}
where $R(f'') = \int [f''(y)]^2 dy$ is the roughness of the second derivative of the underlying density.

The optimal overall bandwidth is obtained by minimizing the MISE. This can be calculated by differentiating \@ref(eq:mise) with respect to $h$ and setting the derivative to zero, yielding
\begin{equation}
  h = \left(\frac{R(K)}{R(f'')n}\right)^{1/5}.
  (\#eq:opth)
\end{equation}
So the optimal $h$ is proportional to $n^{-1/5}$, which clearly satisfies the conditions \@ref(eq:asymptotickde).

However, this value of $h$ depends on the underlying density $f$ which we don't know. For a Gaussian density $g$ with variance $\sigma^2$, $R(g) = \frac{1}{2\sqrt{\pi}}\sigma^{-1}$ and $R(g'') = \frac{3}{8\sqrt{\pi}}\sigma^{-5}$. So if we use a Gaussian kernel, and assume the underlying density has the same roughness as a Gaussian density, we obtain the "normal reference rule":
\begin{equation}
  h = \sigma\left(\frac{4}{3n}\right)^{1/5} = 1.06\sigma n^{-1/5},
  (\#eq:nrr)
\end{equation}
where $\sigma$ is the standard deviation of the underlying density. This is often too large as a Gaussian distribution is relatively smooth (and so has a relatively low $R(f'')$ value). Silverman proposed replacing 1.06 by 0.9 and $\sigma$ by a robust estimate given by $\min(s, IQR/1.34)$, giving his rule-of-thumb \@ref(eq:ruleofthumb).

Another popular bandwidth choice is the Sheather-Jones "plug-in" bandwidth [@SJ91], obtained by replacing $R(f'')$ in \@ref(eq:opth) by an estimate based on the data.  To use this bandwidth, just set `bw = "SJ"` in `geom_density()`.

Bandwidths obtained in this way are designed to give a good overall estimate of the underlying density, but may not be optimal for any particular point of the density. Our goal is to find anomalies in the data, rather than find a good representation for the rest of the data, and so we are interested in the regions of low density.

If we optimized MSE \@ref(eq:mse) rather than MISE \@ref(eq:mise), we would obtain
$$
h = \left(\frac{f(y)R(K)}{n[f''(y)]^2}\right)^{1/5}.
$$
This shows that larger bandwidths are required when $f(y)/[f''(y)]^2$ is relatively large, which occurs in the extreme tails of a distribution. Often the usual bandwidth selection methods result in bandwidths that are too small and can cause the kernel density estimates of observations in the tails to be confused with anomalies. So bandwidths for anomaly detection tend to be a little larger than bandwidths for other purposes.

We can adjust the bandwidth using the argument `adjust` as follows.

```{r oldfaithful5, fig.asp=0.45, fig.cap="Kernel density estimate of Old Faithful eruption durations since 2015, omitting the one eruption that lasted nearly two hours."}
oldfaithful %>%
  filter(duration < 7000) %>%
  ggplot(aes(x = duration)) +
  geom_density(adjust = 2) +
  geom_rug() +
  labs(x = "Duration (seconds)")
```

Here we have doubled the default bandwidth obtained using \@ref(eq:ruleofthumb).

### Boundaries

When a Gaussian kernel is used, a kernel density estimate assumes that the underlying density $f$ is smooth and non-zero on the whole real line. This will cause problems when the true density is actually zero for some regions of the sample space. For example, if all data must be positive, then $f(u)=0$ for $u<0$.

There are modified estimators which deal with this situation, but we won't concern ourselves with them here as this situation will not come up very often in the context of anomaly detection.

## Multivarate kernel density estimation

Suppose our observations are $d$-dimensional vectors, $\bm{y}_1,\dots,\bm{y}_n$. Then the multivariate version of \@ref(eq:kde) is given by [@Scott2015]
\begin{equation}
  \hat{f}(\bm{y}) = \frac{1}{n} \sum_{i=1}^n K_H(\bm{y} - \bm{y}_i),
  (\#eq:mkde)
\end{equation}
where $K_H$ is a multivariate probability density with covariance matrix $\bm{H}$. Whenever we estimate a multivariate kernel density estimate, we will use a multivariate Gaussian kernel given by
$$
  K_H(\bm{u}) = (2\pi)^{-d/2} \text{det}(\bm{H})^{-1/2} \exp \{-\textstyle\frac12 \bm{u}'\bm{H}^{-1}\bm{u} \}.
$$

```{r bivariatebandwidths, include=FALSE}
h1 <- MASS::bandwidth.nrd(of2021$duration)
h2 <- MASS::bandwidth.nrd(of2021$waiting)
```

We will illustrate the idea using a simple bivariate example of 10 observations: the same 10 eruption durations discussed above, along with the corresponding waiting times until the following eruption. These are shown in the figure below along with the contours of bivariate kernels placed over each observation. Here we have used a bivariate Gaussian kernel with bandwidth matrix given by $\bm{H} = \left[\begin{array}{cc}`r round(h1,0)` & 0 \\ 0 & `r round(h2,0)`\end{array}\right]$.

```{r ofdw, dependson="of2021", echo=FALSE, warning=FALSE}
h <- c(13, 300)
k <- expand_grid(
  x = seq(-3 * h[1], 3 * h[1], l = 100),
  y = seq(-3 * h[2], 3 * h[2], l = 100)
) %>%
  mutate(z = dnorm(x, 0, h[1]) * dnorm(y, 0, h[2]))
of2021kde <- of2021 %>%
  mutate(k = list(k)) %>%
  unnest(k) %>%
  mutate(x = x + duration, y = y + waiting)
ggplot() +
  geom_contour(data = of2021kde, aes(x = x, y = y, group = eruption, z = z), bins = 4, col = "gray") +
  geom_point(data = of2021, mapping = aes(x = duration, y = waiting)) +
  labs(x = "Duration (seconds)", y = "Waiting time (seconds)") +
  xlim(147, 287) +
  ylim(3070, 6860)
```

If we add the bivariate kernel functions as in \@ref(eq:mkde), we obtain the bivariate kernel density estimate shown below.

```{r ofbivariate1, dependson="of2021", echo=FALSE}
of2021 %>%
  ggplot(aes(x = duration, y = waiting)) +
  geom_point() +
  geom_density_2d() +
  labs(x = "Duration (seconds)", y = "Waiting time (seconds)") +
  xlim(147, 287) +
  ylim(3070, 6860)
```

Now we will apply the method to the full data set, other than the 2 hour eruption and observations where the subsequent waiting time is more than 2 hours (which are likely to be data errors).

We will use the `geom_density_2d()` function, which by default uses a bivariate Gaussian kernel with diagonal bandwidth matrix where the diagonal values are given by \@ref(eq:nrr).

```{r ofbivariate2, fig.cap="Bivariate kernel density estimate with default bandwidths."}
oldfaithful %>%
  filter(duration < 7000, waiting < 7000) %>%
  ggplot(aes(x = duration, y = waiting)) +
  geom_point(color = "gray") +
  geom_density_2d() +
  labs(x = "Duration (seconds)")
```

Here we see that the short durations tended to be followed by a short waiting time until the next duration, while the long durations tend to be followed by a long waiting time until the next duration. There is one anomalous eruption where a short duration was followed by a long waiting time. The unusual durations between 150 and 180 seconds can be followed by either short or long durations.

If $\bm{H}$ is diagonal with values $h_1,\dots,h_d$, then the estimator is consistent when
$$
  \lim_{n\rightarrow\infty} h = 0
  \qquad\text{and}\qquad
  \lim_{n\rightarrow\infty} nh^d = \infty,
$$
where $h = (h_1h_2\dots,h_d)^{1/d}$ is the geometric mean of the diagonal values of $\bm{H}$. The default values for `geom_density_2d()` satisfy this property.

Note that the diagonal values $h_1,\dots,h_d$ tend to be larger than the values used in the corresponding univariate density estimates, as the convergence properties of the estimate are slower for larger $d$. Consequently, the default bandwidths for `geom_density_2d()` tend to be too small. Further, because we are interested in the tails of the distribution, we usually want larger bandwidths than would be suitable for obtaining good estimates of the density function.

Figure \@ref(fig:ofbivariate3) shows a bivariate kernel density estimate where the bandwidths are double the default values.

```{r ofbivariate3, fig.cap="Bivariate kernel density estimate with double the default bandwidths."}
oldfaithful %>%
  filter(duration < 7000, waiting < 7000) %>%
  ggplot(aes(x = duration, y = waiting)) +
  geom_point(color = "gray") +
  geom_density_2d(adjust = 2) +
  labs(x = "Duration (seconds)")
```
## Extreme value theory {#evt}

Extreme Value Theory is used to model rare, extreme events and is useful in anomaly detection. Suppose we have $n$ independent and identically distributed random variables $Y_1, \dots, Y_n$ with a distribution function $F(y) = P\{Y \leq y\}$. Then the maximum of these $n$ random variables is $M_n = \max \{Y_1, \dots, Y_n\}$. If $F$ is known, the distribution of $M_n$ is given by $P\{M_n \leq z \} = \left(F(z)\right)^n$. However, $F$ is usually not known in practice. This gap is filled by Extreme Value Theory, which studies approximate families of models for $F^n$ so that extremes can be modeled and their uncertainty quantified.

It is well known, due to the central limit theorem, that the average of a set of iid random variables will converge to the normal distribution under relatively weak conditions on the underlying distribution. The Fisher-Tippett-Gnedenko (FTG) Theorem provides an analogous result for the maximum. It was developed in a series of papers by @Frechet1927, @Fisher1928, and @Gnedenko1943. Independently, @Mises1936 proposed a similar result. The FTG Theorem states that if the maximum can be scaled so that it converges, then the scaled maximum will converge to either a Gumbel, Fréchet or Weibull distribution [@coles2001introduction, 46]; no other limits are possible.

### Fisher-Tippett-Gnedenko Theorem {-}

If there exist sequences $\{a_n\}$ and $\{b_n\}$ such that
$$
  P\left\{ \frac{(M_n - a_n)}{b_n} \leq z \right\} \rightarrow G(z) \quad \text{as} \quad n \to \infty,
$$
where $G$ is a non-degenerate distribution function, then $G$ belongs to one of the following families:
\begin{align*}
  &\text{Gumbel} :  && G(z) = \exp\left(-\exp \left[- \Big(\frac{z-b}{a}\Big) \right] \right), \quad -\infty < z < \infty , \\
  &\text{Fréchet} : && G(z) =
    \begin{cases}
      0 ,                                                           & z \leq b  , \\
      \exp \left( - \left( \frac{z-b}{a}\right)^{-\alpha} \right) , & z > b    ,
    \end{cases}                     \\
  &\text{Weibull} : && G(z) =
    \begin{cases}
      \exp \left( - \left(- \left[\frac{z-b}{a}\right]\right)^{\alpha} \right) , & z < b  ,    \\
      1 ,                                                                        & z \geq b  ,
    \end{cases}
\end{align*}
for parameters $a, b$ and $\alpha$ where $a, \alpha >0$.

These three families of distributions can be further combined into a single family by using the following distribution function known as the Generalized Extreme Value (GEV) distribution,
\begin{equation}
  G(z) = \exp\left\{ -\left[ 1 + \xi\Big(\frac{z - \mu}{\sigma} \Big)\right]^{-1/\xi} \right\} ,
  (\#eq:EVT4)
\end{equation}
where the domain of the function is $\{z: 1 + \xi (z - \mu)/\sigma >0 \}$. The location parameter is $\mu\in\mathbb{R}$, $\sigma>0$ is the scale parameter, while $\xi\in\mathbb{R}$ is the shape parameter. When $\xi = 0$ we obtain a Gumbel distribution with exponentially decaying tails. When $\xi < 0$ we get a Weibull distribution with a finite upper end, and when $\xi > 0$ we get a Fréchet family of distributions with heavy tails including polynomial tails.

If we take the negative of the random variables $Y_1,\dots,Y_n$, it becomes clear that a similar result holds for the minimum.

The three types of limits correspond to different forms of the tail behaviour of $F$.

  * When $F$ has a finite upper bound, such as with a uniform distribution, then $G$ is a Weibull distribution.
  * When $F$ has exponential tails, such as with a normal distribution or an exponential distribution, then $G$ is a Gumbel distribution.
  * When $F$ has heavy tails including polynomial decay, then $G$ is a Fréchet distribution. One example is when $F$ itself is a Fréchet distribution with $F(y)= e^{-1/y}$, $y>0$.

To illustrate, suppose $F$ is a standard normal distribution N(0,1) and we have $n=1000$ observations. Then the distribution of the maximum can be obtained via simulation. We will fit both a kernel density estimate and an extreme value distribution to the resulting maximums obtained.

```{r evdexample, fig.cap="Distribution of the maximum of 1000 N(0,1) draws. Here we have simulated 10000 such maximums and shown a kernel density estimate along with the density from an estimated GEV distribution."}
m <- 10000 # Number of simulations
n <- 1000 # Number of observations in each data set
maximums <- numeric(m)
for (i in seq(m)) {
  maximums[i] <- max(rnorm(n))
}
gev <- evd::fgev(maximums)$estimate
truedensity <- tibble(y = seq(2, 6, l = 100)) %>%
  mutate(fy = evd::dgev(y, loc = gev["loc"], scale = gev["scale"], shape = gev["shape"]))
tibble(maximum = maximums) %>%
  ggplot(aes(x = maximums)) +
  geom_density() +
  geom_line(data = truedensity, aes(x = y, y = fy), col = "red")
```

### The Generalized Pareto Distribution {-}

The Peaks Over Threshold (POT) approach regards extremes as observations greater than a threshold $u$. The probability distribution of *exceedances* above a specified threshold $u$ can be expressed as
\begin{equation}
  H(y) = P\left \{Y \leq u + y \mid Y > u \right \}
  = \frac{ F(u+y) - F(u)}{1 - F(u)}. (\#eq:POT3)
\end{equation}
When the distribution $F$ satisfies the FTG theorem, then [@coles2001introduction, 75] $H$ is a  **Generalized Pareto Distribution** (GPD) defined by
  \begin{equation}
    H(y) \approx 1 - \Big( 1 + \frac{\xi y}{\sigma_u} \Big)^{-1/\xi} ,
    (\#eq:POT)
  \end{equation}
where the domain of $H$ is $\{y: y >0 \text{ and } (1 + \xi y)/\sigma_u >0 \}$, and $\sigma_u = \sigma + \xi(u- \mu)$. The GPD parameters are determined from the associated GEV parameters. In particular, the shape parameter $\xi$ is the same in both distributions.

Continuing the previous example, we now look at the probability distribution of exceedances above 3 from a N(0,1) distribution. We simulate 1 million N(0,1) values and only keep those above 3. Then a kernel density estimate and two GPD estimates are drawn. The red GPD uses the parameters obtained previously from the GEV estimate, while the blue GPD estimates the parameters from the exceedances.

```{r pot, dependson="evdexample", fig.cap="Conditional distribution of exceedances above 3 from N(0,1) draws. Here we have simulated 1 million values and only kept those above 3. A kernel density estimate is shown (in black) along with the density implied by the GEV distribution (in red), and an estimated GPD distribution (in blue). The boundary at 3 causes the kernel density estimate to be biased around 3."}
df <- tibble(y = rnorm(1e6)) %>%
  filter(y > 3)
gpd <- evd::fpot(df$y, 3)$estimate
truedensity <- truedensity %>%
  mutate(
    hy = evd::dgpd(y, loc = 3, scale = gev["scale"], shape = gev["shape"]),
    hy2 = evd::dgpd(y, loc = 3, scale = gpd["scale"], shape = gpd["shape"])
  )
df %>%
  ggplot(aes(x = y)) +
  geom_density() +
  geom_line(data = truedensity, aes(x = y, y = hy), col = "red") +
  geom_line(data = truedensity, aes(x = y, y = hy2), col = "blue")
```

The red estimate is better because it is based on more information (10000 maximums rather than `r NROW(df)` exceedances). The kernel density estimate (in black) is biased around 3 because of the boundary problem discussed in Section \@ref(kde). Otherwise, the kernel density estimate closely matches the GPD estimates.
