# Univariate methods

In this chapter, we will explore various methods for identifying and displaying anomalies when there is only one variable.  Throughout the chapter, we will use three illustrative examples.

1. The cricket batting averages discussed in Section \@ref(sec:scatterplots). Recall that this appeared to contain one genuine outlier, the batsman Don Bradman.
2. Old Faithful eruption durations since 2015, also discussed in Section \@ref(sec:scatterplots). For these data, there were two extreme anomalies --- one at nearly 2 hours and one at 1 second --- and several "inliers" in a area of low density between 140 and 180 seconds.
3. One thousand draws from a N(0,1) distribution. Of course, this contains no genuine outliers, as all the data come from the same underlying probability distribution. This example is used to check how many false anomalies different methods identify.

```{r n01, fig.asp=0.2, fig.cap="One thousand observations from a N(0,1) distribution."}
set.seed(1)
n01 <- tibble(y = rnorm(1000))
n01 %>%
  ggplot(aes(x=y, y=1)) + 
  geom_jitter(width=0, alpha=0.5) +
  scale_y_discrete() + 
  labs(y="", x="N(0,1)")
``````

## Z scores and related tests

Let's start with some bad ideas that are still sometimes used. Probably the most widely used is the idea of $z$-scores which assume the data come from a normal distribution. If our data are given by $y_1,\dots,y_n$, then their $z$ scores are given by
$$z_i = (y_i - \bar{y})/s_y$$
where $\bar{y}$ is the mean and $s_y$ is the standard deviation of the observations.

Some books recommended that observations where the absolute value of $z$ is above some threshold (usually 3) be identified as anomalies. This is a bad idea for several reasons.

1. If the data really do come from a normal distribution that has been contaminated with anomalies, then the estimated mean ($\bar{y}$) and standard deviation $s_y$ will also be affected by those anomalies. Any anomaly detection method should be relatively robust to the anomalies in the data.
2. Having a fixed threshold regardless of sample size means that the probability of a spurious anomaly increases with the sample size. 
3. The assumption of normality is very unlikely to be satisfied with most real data. 

Let's see what happens when we apply this test to the three examples.

```{r zscore1}
cricket_batting %>%
  filter(Innings > 20) %>%
  mutate(z = (Average - mean(Average))/sd(Average)) %>%
  filter(abs(z) > 3) %>%
  select(Player, Average, z)

oldfaithful %>%
  mutate(z = (duration - mean(duration, na.rm=TRUE))/sd(duration, na.rm=TRUE)) %>%
  filter(abs(z) > 3)

n01 %>%
  mutate(z = (y - mean(y))/sd(y)) %>%
  filter(abs(z) > 3)
```

 1. Don Bradman is correctly identified as the only outlier in the cricket batting averages.
 2. Only the extreme 2-hour duration is identified. We already know there was an additional strange 1-second eruption, plus several unusual eruptions between 140 and 180 seconds in length. 
 3. One spurious anomaly is identified in the N(0,1) observations.
 
If we replace the mean and standard deviation by more robust estimates (the median and IQR/1.34), we obtain the same results in the first and last examples, but even worse results in the second example.

```{r zscore2}
oldfaithful %>%
  mutate(z = (duration - median(duration, na.rm=TRUE))/
                (IQR(duration, na.rm=TRUE)/1.34)) %>%
  filter(abs(z) > 3)
```

Now there are 209 out of `r NROW(oldfaithful)` observations identified as anomalous, including all of those below 5 minutes. 

In any case, even if the data did come from a normal distribution, the probability of an observation being more than 3 standard deviations from the mean is `r round(2*pnorm(-3),4)`, so we would expect to see 1 in every `r round(.5/pnorm(-3))` regular observations being identified as an anomaly using this approach.

### Peirce's criterion {-}

Benjamin Peirce was a Harvard mathematician in the mid 1800s who worked with astronomical data which were prone to anomalous observations. In @Peirce1852 he proposed the first known test based on what are now called $z$-scores. His criteria was that any observations with $|z| > c$ should be "rejected", where $c$ is a complicated function depending on the sample size $n$ and the number of suspected anomalies. Figure \@ref(fig:zscorecriteria) shows the value of $c$ as a function of sample size, when there is only one suspected anomaly.

```{r zscorecriteria, echo=FALSE, fig.cap="Criteria for anomalies based on z-scores."}
n <- exp(seq(log(3), log(1e5), l=100))
threshold <- numeric(length(n))
for(i in seq_along(n))
  threshold[i] <- weird:::peirce_threshold(n[i])
tibble(n=n, c=threshold) %>%
  ggplot(aes(x=n, y=c)) +
    geom_line() +
    geom_line(aes(y = qnorm(1-0.25/n)), col='red') +
    scale_x_log10() +
    labs(x = "n (sample size)", y = "c (threshold)") +
    geom_label(aes(x=1500, y=4.1, label="Peirce's criterion")) +
    geom_label(aes(x=1e4,  y=3.55, label="Chauvenet's criterion"), col='red')
```

The threshold increases with sample size $n$ to allow for the increasing likelihood of observations from the extreme tails of the distribution.

The `peirce_anomalies()` function returns a logical vector indicating which observations are anomalous under this criterion. Let's apply it to the three examples.

```{r peirce_examples}
cricket_batting %>% filter(peirce_anomalies(Average))
oldfaithful %>% filter(peirce_anomalies(duration))
n01 %>% filter(peirce_anomalies(y))
```

1. When applied to the test cricket batting averages, it doesn't even find the obvious anomaly of Don Bradman.
2. When applied to the Old Faithful eruption durations, it finds only the most extreme duration.
3. Again, one spurious anomaly is identified in the N(0,1) observations.

This failure is at least partly due to the standard error being highly inflated by the existence of the anomalies in the data set. 

### Chauvenet's criterion {-}

Peirce's proposal was largely superseded by an alternative proposed by the astrophysicist William Cauvenet, which was much simpler to describe. He suggested [@Chauvenet1863] replacing the threshold $c$ by the $1-1/(4n)$ quantile from the standard normal distribution, so that any observations with $|z_i| > \Phi^{-1}\left(1-\frac{1}{4n}\right)$ are identified as anomalies. This threshold is also shown in Figure \@ref(fig:zscorecriteria). A consequence of this choice is that Chauvenet's criterion will reject, on average, half an observation of genuine data from a normal distribution regardless of the value of $n$.

The `chauvenet_anomalies()` function can be used to implement this test. For our three examples, it gives identical results to those above for `peirce_anomalies()`.

### Maximum z-score test {-}

By the 20th century, the concept of hypothesis testing had been developed and was applied to the identification of anomalies. @Thompson1935 proposed a test using extreme $z$-scores, where an observation is considered an anomaly if
$$
\max_i |z_i| > c_\alpha
$$
where $c_\alpha$ is a threshold that depends on the size of the test $\alpha$. The idea was quickly picked up by Egon Pearson and Chandra Sekar who proposed [@Pearson1936] setting
$$
  c_\alpha = t_{\alpha/n, n-2} \sqrt{\frac{n-1}{n-2+t^2_{\alpha/n, n-2}}},
$$
where $t_{\alpha/n, n-2}$ is the $1-\alpha/n$ quantile of the $t$ distribution with $n-2$ degrees of freedom. Later, this was extended by [@Grubbs1950] to handle multiple outliers, and the test is often now called "Grubbs' test".


$$
\max_i |z_i| > \frac{n-1}{\sqrt{n}}\sqrt{\frac{t^2_{\alpha/(n(n-1)), n-2}}{n-2 + t^2_{\alpha/(n(n-1)), n-2}}}
$$
where $t_{\alpha/(2n), n-2}$ is the $1-\alpha/(2n)$ quantile of the $t$ distribution with $n-2$ degrees of freedom.

The test was meant to be applied iteratively, where observations were removed from the data if determined to be anomalies, and the test re-applied to the remaining data. This process continues until no more anomalies are found.

We can apply the test using the `grubbs.test()` function from the `outliers` package.

```{r}
library(outliers)
grubbs.test(cricket_batting %>%
              filter(Innings > 20) %>%
              pull(Average)
)
```

## Statistical tests for univariate anomalies

### Dixon's Q test

If $y_{(1)},\dots,y_{(n)}$ denote the ordered values of our sample, then Dixon's Q statistic [@Dixon1950] is given by
$$
  Q = \frac{y_{(n)} - y_{(n-1)}}{y_{(n)} - y_{(1)}},
$$
the ratio of the difference between the two largest values to the range of the data. A variant of the statistic uses $y_{(2)}-y_{(1)}$ in the denominator instead, the difference between the second smallest and minimum observations. For samples sizes $n$ in the range 3 to 30, we can use the `dixon.test()` function from the `outliers` package to perform the test. For larger sample sizes, critical values are not available for this test.

## Boxplots

Boxplots were invented by John Tukey as a quick summary of medium sized data sets. They are widely used to identify anomalies, which are shown as separate points in the plot.

Figure \@ref(fig:cricketbox) shows a boxplot of the cricket batting average data, previously shown in  \@ref(fig:cricket1).

```{r cricketbox, fig.asp=0.2, fig.cap="(ref:cricket1)"}
cricket_batting %>%
  filter(Innings > 20) %>%
  ggplot(aes(x=Average, y=1)) +
  geom_boxplot() +
  scale_y_discrete() +
  labs(y="", x="Career batting average")
```

The middle line in the box shows the median, and the ends of the box are the fourths ($L_F$ and $U_F$). Half of all observations lie within the box. The width of the box is called the "interquartile range", defined by $IQR = U_F - L_F$. Any points more than $1.5*IQR$ outside the box are shown as anomalies and appear as separate points in the plot. The "whiskers" that extend out each side of the box show the range of the remaining points.

Originally, Tukey proposed two levels of outliers --- those more than $1.5*IQR$ beyond the box were labelled "outside" values, while those more than $3*IQR$ beyond the box were labelled "far out" values. Most software implementations of boxplots do not distinguish between these groups.

In this cricket batting example, the boxplot works well because the data set is not too large or small, and the distribution of points other than the anomaly is unimodal.

However, boxplots can be misleading, and are they are limited in at least two respects.

1. For large data sets, boxplots show too many points as anomalies, and it is hard to distinguish them.
2. Boxplots assume that the distribution of the data is unimodal.

```{r standardnormal, echo=FALSE}
q1 <- qnorm(0.25)
q3 <- qnorm(0.75)
iqr <- q3 - q1
whisker1 <- q1 - 1.5*iqr
whisker2 <- q3 + 1.5*iqr
```

To better understand the first problem, imagine the data comprised $n$ observations from a standard normal distribution N(0,1). Figure \@ref(fig:normalboxplot) shows an example with 10000 points.

```{r normalboxplot, fig.asp=0.2, fig.cap="Boxplot of 10000 draws from a standard normal distribution."}
tibble(x = rnorm(10000)) %>%
  ggplot(aes(x=x, y=1)) +
  geom_boxplot() +
  scale_y_discrete() +
  labs(y="")
```

Many anomalies are shown, but since all observations come from a simple distribution, none of them are actually anomalies. For this distribution, Q1 $= `r round(q1,2)`$, Q3 $= `r round(q3,2)`$, IQR $= `r round(iqr,2)`$, and so any observations less than $`r round(q1,2)` - 1.5\times `r round(iqr,2)` = `r round(whisker1, 2)`$ or greater than $`r round(q3,2)` + 1.5\times `r round(iqr,2)` = `r round(whisker2, 2)`$ would be identified as "anomalous" by a boxplot and plotted as separate points. The probability of a standard normal observation being at least $`r round(whisker2, 2)`$ in absolute value is $`r sprintf("%.5f", 1-pnorm(whisker2))`$. So with 10000 observations, we would have $`r round(10000*(1-pnorm(whisker2)))`$ anomalies identified, none of which would be a genuine anomaly.

The second problem is demonstrated using the Old Faithful eruption duration data, shown in Figure \@ref(fig:oldfaithful1). As before, we will omit the largest value so we can see the details in the remaining data.

```{r oldfaithful2a, fig.asp=0.2, fig.cap="Boxplot of Old Faithful eruption durations since 2015, omitting the one eruption that lasted nearly two hours."}
oldfaithful %>%
  filter(duration < 6000) %>%
  ggplot(aes(x=duration, y=1)) +
  geom_boxplot() +
  labs(y="", x="Duration (seconds)") +
  scale_y_discrete()
```

Quite a few anomalies are shown, including the one-second eruption we identified earlier. But the remaining "anomalies" are not particularly unusual observations. All the points below 180 seconds are identified as anomalies, even though we know that observations in the region between 100 and 140 are not unusual for this geyser. Because the boxplot does not allow for more than one mode, all the points in the second smaller cluster are identified as anomalies. The points around 300 seconds are also not really anomalies --- these are just values in the upper tail of the distribution for eruptions.

## Letter value plots

The problem that boxplots have with large data sets was addressed by @Hofmann2017-hz who introduced "letter-value" plots, a variation of boxplots that replace the whiskers with a variable number of letter values. In these plots, each pair of letter values marks the boundaries of a box. The box bounded by the fourths is the same as the box of a boxplot; the additional boxes extend 

These can be produced using the `lvplot` package.

```{r cricketboxlv, fig.asp=0.2, fig.cap="Letter value plot of career batting averages for all men and women who played test cricket and batter more than 20 times."}
library(lvplot)
cricket_batting %>%
  filter(Innings > 20) %>%
  ggplot(aes(x=1,y=Average)) +
  geom_lv(aes(fill=..LV..)) +
  scale_x_discrete() +
  coord_flip() +
  labs(x="", y="Career batting average") +
  theme(legend.key.height = unit(.2,"cm")) 
```

```{r lv, echo=FALSE}
lv <- cricket_batting %>%
  filter(Innings > 20) %>%
  stat_lv(mapping = aes(y=Average))
```

Here the median is given by M, the fourths by F, and so on. So the middle box (F) is bounded by the fourths and contains all but 2/4 the data; the next box (E) is bounded by the eighths and contains all but 2/8 of the data; then D is bounded by the sixteenths and contains all but 2/16 of the data; and so on. 

In this example, the most extreme box (labelled Z) is bounded by the points that fall within the 1/256 letter values. So it contains all but 2/256 of the data, and shows $2 / 256 \times `r NROW(filter(cricket_batting, Innings > 20))` = `r round(2/256*NROW(filter(cricket_batting, Innings > 20)))`$ points as anomalies. 

The stopping rule used in the letter value plot is to show the boxes up to letter value $k$, where
$$
  0.5\sqrt{2d_k} z_{1-\alpha/2} > d_{k+1}
$$
and $z_{1-\alpha/2}$ is the $1-\alpha/2$ quantile of a standard Gaussian distribution. This choice is based on the idea that the edges of the boxes are quantile estimates, and the confidence interval for each  quantile estimate that is displayed should not overlap the subsequent quantile estimate. The Gaussian distribution arises because the quantile estimate has an approximate Gaussian distribution due to the Central Limit Theorem. This stopping rule means that, on average, there should be fewer than $2z^2_{1-\alpha/2}$ legitimate observations in the tails. By default, $\alpha=0.05$, so that, on average, there should be fewer than $2 \times (`r qnorm(1-0.05/2)`)^2 = `r round(2* qnorm(1-0.05/2)^2,1)`$ legitimate observations in the tails, regardless of the size of the data set.

In this cricketing example, it looks like there is one true anomaly (Don Bradman) and the remaining 8 identified "anomalies" are simply in the tails of the distribution of the remaining data.

## HDR boxplots

## Lookout

Denote the kernel density estimate of $y_i$ by $f_i$ and the leave-one-out kde of $y_i$ (by leaving out $y_i$) by $f_{-i}$. Then we fit a Generalized Pareto Distribution to $-\log(f_i)$ using the POT approach discussed in Section \@ref(sec:evt). We use the $90^{\text{th}}$ percentile as the threshold for the POT approach. Using the fitted GPD parameters, $\mu$, $\sigma$ and $\xi$, we declare points with $P\left(-\log(f_{-i})|\mu, \sigma,\xi \right) < \alpha$ to be outliers. 

The output probability of lookout is the GPD probability of the points, so that low probabilities indicate likely outliers and high probabilities indicate normal points. 

